---
title: "p8105_hw2_jr4550"
author: Jeff Remo 
output: github_document
date: "2025-10-01"
---


```{r}
library(tidyverse)
```

## Problem 1
First, clean the data in pols-month.csv.
```{r}
pols_df = 
  read_csv("data/pols-month.csv", na = c("NA", ".", "")) %>% 
  janitor::clean_names() %>% 
  separate(mon, into = c("year", "month", "date"), sep = "-", convert = TRUE) %>% 
  mutate(
    month = month.name[month], 
    president = case_when(
      prez_gop == 1 ~ "gop",
      prez_dem == 1 ~ "dem"
    )) %>% 
  select(-prez_gop, -prez_dem)
  
```
Cleaning snp.csv dataset 

```{r}
snp_df = 
  read_csv("data/snp.csv", na = c("NA", ".", "")) %>% 
  janitor::clean_names() %>%
  separate(date, into = c("month", "day", "year"), sep = "/", convert = TRUE) %>%
  mutate(
    month = month.name[month],
    year = if_else(year < 50, year + 2000, year + 1900)
  ) %>%
  select(year, month, close)
```

Cleaning unemployment.csv dataset 
```{r}
unemployment_df = 
  read_csv("data/unemployment.csv", na = c("NA", ".","")) %>% 
  janitor::clean_names() %>% 
   pivot_longer(
    cols = jan:dec,
    names_to = "month",
    values_to = "unemployment_pct"
  ) |>
  mutate(
    year = as.integer(year),
    month = factor(month,
                   levels = tolower(month.abb),
                   labels = month.name),
    month = as.character(month)
  )
```
Joining data sets
```{r}
merged_df = pols_df %>%
  left_join(snp_df, by = c("year", "month")) %>%
  left_join(unemployment_df, by = c("year", "month")) %>% 
  select(-date)
```

#### Description of Problem 1 datasets
The pols_df dataset characterizes the the political party representation of the number of Senators (sen_dem & sen_gop) and Representatives (rep_gop & rep_dem), and the political party of the President (dem/gop) in the United States from 1947-2015. The snp_df shows the S&P values of each month from Dec 1950- July 2015. The unemployment_df shows the unemployments rates in the US from Jan 1948-June2015.


## Problem 2
Read and clean Mr.Trashwheel Dataset

```{r}
library(tidyverse)
library(readxl)

mr_trashwheeldf = 
  read_excel("data/202509 Trash Wheel Collection Data.xlsx", 
             sheet = "Mr. Trash Wheel",
             range = "A2:N586") %>%
  janitor::clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel",
    year = as.numeric(year)
  )

prof_trashwheel_df = 
  read_excel("data/202509 Trash Wheel Collection Data.xlsx", 
             sheet = "Professor Trash Wheel",
             range = "A2:M108") %>%
  janitor::clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(
    trash_wheel = "Professor Trash Wheel",
    year = as.numeric(year)
  )

gwynnda_df = 
  read_excel("data/202509 Trash Wheel Collection Data.xlsx", 
             sheet = "Gwynns Falls Trash Wheel",
             range = "A2:L157") %>%
  janitor::clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(
    trash_wheel = "Gwynns Falls Trash Wheel",
    year = as.numeric(year)
  )

trash_wheel_df = 
  bind_rows(mr_trashwheeldf, prof_trashwheel_df, gwynnda_df)

prof_total_weight = prof_trashwheel_df %>%
  summarize(total_weight = sum(weight_tons, na.rm = TRUE)) %>%
  pull(total_weight)

gwynnda_june_2022_cigs = gwynnda_df %>%
  filter(month == "June", year == 2022) %>%
  summarize(total_cigs = sum(cigarette_butts, na.rm = TRUE)) %>%
  pull(total_cigs)
```

```{r}
format(prof_total_weight, nsmall = 2)
format(gwynnda_june_2022_cigs, big.mark = ",")
```

#### Description of Trash Wheel Dataset 

The combined Trash Wheel dataset contains r nrow(trashwheel_df) observations from three water-wheel trash collectors in Baltimore's Inner Harbor. Key variables include date, weight_tons, volume_cubic_yards, and counts of specific items like plastic_bottles, cigarette_butts, and sports_balls. Professor Trash Wheel collected a total of 216.26 tons of trash. In June 2022, Gwynnda collected 18,120 cigarette butts.

## Problem 3 
Zillow and Zipcodes datasets 

#### Zipcode dataset
```{r}
zip_codes_df = 
  read_csv("data/Zip Codes.csv", na = c("NA", ".", "")) %>%
  janitor::clean_names() %>%
  rename(zipcode = zip_code) %>%
  select(zipcode, county, neighborhood) %>%
  # Remove duplicates - keep first occurrence of each zipcode
  distinct(zipcode, .keep_all = TRUE)
```
#### Zillow dataset 
```{r}
zillow_df = 
  read_csv("data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv", na = c("NA", ".", "")) %>%
  janitor::clean_names() %>%
  rename(zipcode = region_name) %>%
  select(zipcode, county_name, starts_with("x2")) %>%
  pivot_longer(
    cols = starts_with("x2"),
    names_to = "date",
    values_to = "rental_price"
  ) %>%
  mutate(
    date = str_remove(date, "^x"),
    date = as.Date(date, format = "%Y_%m_%d")
  )
```
#### combine datasets 
```{r}
rental_df = 
  zillow_df %>%
  left_join(zip_codes_df, by = "zipcode") %>%
  rename(borough = county_name) %>%
  select(zipcode, borough, neighborhood, date, rental_price) %>%
  arrange(zipcode, date)
```

#### Description of dataset 
```{r}
nrow(rental_df)
rental_df %>% distinct(zipcode) %>% nrow()
rental_df %>% distinct(neighborhood) %>% filter(!is.na(neighborhood)) %>% nrow()
```
Number of observations: 17284
Number of Unique Zipcode: 149
Number of Unique Neighborhoods: 42

#### Zipcodes in zip_codes_df vs zillow_df
```{r}
missing_zips = zip_codes_df %>%
  anti_join(zillow_df %>% distinct(zipcode), by = "zipcode") %>%
  arrange(zipcode)
n_missing = nrow(missing_zips)
```

Zipcodes in "missing_zips" show the ZIP codes in zip_codes_df that are not in the zillow_df. This could be for a few reasons such as some zipcodes may not have any or have few Zillow rentals. One example is the Empire State Building, as it has its own zipcode, 10118. 

#### Covid Price Comparison
```{r}
  price_comparison = 
  rental_df %>%
  filter(date %in% as.Date(c("2020-01-31", "2021-01-31"))) %>%
  select(zipcode, borough, neighborhood, date, rental_price) %>%
  distinct(zipcode, date, .keep_all = TRUE) %>%
  pivot_wider(
    names_from = date,
    values_from = rental_price
  ) %>%
  rename(
    jan_2020 = `2020-01-31`,
    jan_2021 = `2021-01-31`
  ) %>%
  filter(!is.na(jan_2020), !is.na(jan_2021)) %>%
  mutate(
    price_change = jan_2021 - jan_2020,
    pct_change = (price_change / jan_2020) * 100
  ) %>%
  arrange(price_change) %>%
  slice_head(n = 10)
```

```{r}
price_comparison %>%
  select(zipcode, borough, neighborhood, jan_2020, jan_2021, price_change, pct_change) %>%
  knitr::kable(
    col.names = c("ZIP Code", "Borough", "Neighborhood", "Jan 2020", "Jan 2021", 
                  "$ Change", "% Change"),
    digits = 2,
    format.args = list(big.mark = ",")
  )
```
##### Comment: 
The 10007 ZIP code (New York County, Lower Manhattan) saw the largest drop in rental prices. In January 2020, the average rent was $6,334.21, which fell to $5,421.61 in January 2021, a decrease of $912.60, or 14.41%.

